---
title: 'CellSpectra: Advanced Parameters and Driver Genes'
author: "Konstantin A. Klötzer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CellSpectra parameters and advanced usage
We will now focus on some of the CellSpectra parameters and how they impact 
analysis results. We will first load our packages, define an output_folder_base, 
load a seurat object (subsetted version of our extended human kidney atlas), and
define some columns for seurat_object preparation (see our Query_Reference_
Tutorial if not familiar).

Note: This object has condition information stored in "disease_grouped_qr" column
with one query sample ("sample_of_interest") and a few healthy reference samples

```{r Load and prepare, message=FALSE, warning=FALSE}
library(Matrix)
library(dplyr)
library(Seurat)
library(CellSpectra)
library(KEGGREST)

output_folder_base = "output/" 

seurat_object = readRDS("Human_PT_subset.rds")

# We assume the respective metadata is stored in the following columns
sample_id_col <- "orig_ident"
condition_col <- "disease_grouped_qr"
celltype_annotation_col <- "C_scANVI_modified1"

#we use the prepare function to get our seurat object in shape
seurat_object <- prepare_seurat_object(
  seurat_obj = seurat_object, #our object to prepare
  celltype_col = celltype_annotation_col, #column in which annotation is stored 
  sample_id_col = sample_id_col, #sample identifier
  condition_col = condition_col, #column storing the condition information
  query_list = c("sample_of_interest"), #categories in the condition_col defining the query samples 
  control_list = c("reference_samples") #categories in the condition_col defining the reference
)
```
## Check the number of cells in our object
To keep things easy, our provided object only contains PT cells. We will now
check the cell numbers per sample. Consider that sample 32-2 is our diseased
patient. 

```{r Check Cell numbers, echo=TRUE}
table(seurat_object$orig_ident)
```

As we can see, 2 of our reference samples have 1.5-4 times the cells of our 
query. Since scRNA-seq data is sparse, different cell numbers / capturing depth
could introduce technical noise. CellSpectra controls these differences by 
matching the total counts to the counts of our query. To not lose too many cells
we could create pseudo-replicates per reference sample. 

Lets assume our query has 300 cells, and most reference samples have many more 
(e.g. 3000 cells) we could make approx. 10 reference pseudoreplicates from one
reference sample. In other words, we would lose cells. Here 2 or 3 replicates 
would seem reasonable. However, we don't recommend using pseudo-replicates if 
1. Cell numbers of the query are similar or even larger compared to the reference 
samples and 
2. Total number of biological replicates in the reference is small. 
Using pseudoreplicates would result in false homogeneity in the LOO approach. 
In the manuscript, since we were usually working with many 
reference samples (50 in the human extended kidney), we used 3 pseudoreplicates. 

Here, due to the small number of reference samples, we will not use 
pseudoreplicates for now.

## Create on_the_fly references and create a KEGG gene set database
Lets create the reference (and a Kegg gene set database if not already provided)
```{r Create Reference, echo=TRUE, message=FALSE, warning=FALSE}
#Step 1
create_references(
  seurat_object = seurat_object,
  output_folder_base = output_folder_base,
  num_replicates = 1, 
  cell_types = c("PT"), 
  cell_number_threshold = 10,
  seed = 123
)

#this could take a while to run - we provide a computed version
#kegg_gene_matrix <- process_gene_sets_from_KEGG("hsa", seurat_object = seurat_object)
#saveRDS(kegg_gene_matrix, "output/go_sets_modified.rds")
```

We should now have the folder output/PT and the file samples_of_interest.rds. 
In output/PT/ we should find a subfolder for our query sample (storing the query
and on-the-fly reference gene expression matrix) and a valid_samples file (this 
file would store information which samples had enough cells included to be
considered). 

This is everything we need to run CellSpectra. Consider, that we will not filter
any low expression genes for now. 

```{r Run Spectra, echo=TRUE, message=FALSE, warning=FALSE}
#Step 2
run_spectra(output_folder_base = output_folder_base, cell_types = c("PT"), CHISQ.MAX = 4, expression_threshold = -1, gene_number_threshold = 10, QC_report = TRUE)
```
We should now have three additional subfolders of output/

1. R2: Here we store R2 values for all gene set of the query compared to the reference
2. Pval: Pvalus of our sample level Chi-Square-Statistic of residuals 
3. Padj: Same Pvals after correcting for multiple testing (fdr) for several pathways

Lets open the R2 file: 

```{r Loading results, echo=TRUE, message=FALSE, warning=FALSE}
#load results
r2_df = read.csv("output/R2/R2_PT.csv", row.names = 1)
r2_df = t(r2_df)
```

In this case, the KEGG pathway "Staphylococcus.aureus.infection" had the lowest 
R2 of ~0.24. However, while this indicates that the query sample gene expression pattern
could not be predicted based on the reference expression pattern, it doesn't tell
us if this is expected inter-sample variance (= humans are not coordinated in 
this function) or a significant deviation from a usually coordinated gene expression 
pattern (query is "dyscoordinated"). It is also possible, that technical noise 
is responsible for the low degree of coordination. To better understand what is 
happening, we will now take a look at our QC report: 

```{r Load QC, echo=TRUE, message=FALSE, warning=FALSE}
#Step 2
qc_report_query = read.csv("output/PT/subfolder_sample_32-2/pathway_QC.csv", row.names = 1)

```

Our three reference samples showed a mean R2 0.55 with a SD of 0.28 and 
a V1 Variance explained of 0.89. This indicates, that within the reference
the estimation of V1 was already poor and random. Reference samples did not show
a convincing pattern of coordination. However, we will now further consider the 
pvalue of our statistical test to provide more confidence.

```{r Loading pval, echo=TRUE, message=FALSE, warning=FALSE}
#load results
pval_df = read.csv("output/Padj/Padj_PT.csv", row.names = 1)
pval_df = t(pval_df)
```

As we check the pvalues for all gene sets, we see that many pathways were significant
after fdr correction. We generally assume that the real representation of biological 
inter-sample variance is poor, since we only are working with 3 different 
reference samples. Lets focus on the cAMP.signaling.pathway, which showed a high
ranking based on the sample pvalue:

R2 of 0.91 with a LOO mean R2 of 0.94, SD of 0.02 and a V1VE of 0.98. 

This is a highly coordinated pathway within our reference. While our query R2 of
0.91 is still high, our p value gives us confidence that our diseased sample shows
significant changes within this pathway. 

## Focusing on driver genes
We next want to understand which gene expression changes are responsible for the 
observed "dyscoordination" in the cAMP.signaling.pathway. We can use our 
get_driver_genes() function to do that: 
```{r Get Driver Genes, echo=TRUE, message=FALSE, warning=FALSE}

CHISQ.MAX = 4
gene_set_name = "hsa04024 cAMP signaling pathway - Homo sapiens (human)"

drivinggenes1 <- get_driver_genes(output_folder_base, gene_set_name, 
                                 cell_type = "PT", 
                                 sample_of_interest = "32-2")

```

```{r Plot Driver Genes, echo=TRUE, message=FALSE, warning=FALSE}

#we can load our query and on-the-fly reference for plotting
datH = readRDS("output/PT/subfolder_sample_32-2/datH.rds")
datD = readRDS("output/PT/subfolder_sample_32-2/datD.rds")

#positive driver gene example
gene_neg <- "CREB3L2"
expression_H_neg <- datH[, gene_neg]
expression_D_neg <- datD[, gene_neg]

# Calculate density
density_H_neg <- density(expression_H_neg)

# Set x-axis and y-axis to start from 0, and add a slight margin to max values
xlim_neg <- c(0, max(density_H_neg$x, expression_D_neg) * 1.05)
ylim_neg <- c(0, max(density_H_neg$y) * 1.05)

# Plot the density with thicker blue line, ensuring axes meet at 0 precisely
plot(density_H_neg, main = paste(gene_neg), 
     xlab = "", ylab = "", col = "blue",
     xlim = xlim_neg, ylim = ylim_neg, lwd = 4, cex.axis = 2,
     xaxs = "i", yaxs = "i")

# Add red dashed line and points
abline(v = expression_D_neg, col = "red", lwd = 4, lty = 2)
points(expression_D_neg, 0, col = "red", pch = 19)

#negative driver gene example
gene_neg <- "CNGA1"
expression_H_neg <- datH[, gene_neg]
expression_D_neg <- datD[, gene_neg]

# Calculate density
density_H_neg <- density(expression_H_neg)

# Set x-axis and y-axis to start from 0, and add a slight margin to max values
xlim_neg <- c(0, max(density_H_neg$x, expression_D_neg) * 1.05)
ylim_neg <- c(0, max(density_H_neg$y) * 1.05)

# Plot the density with thicker blue line, ensuring axes meet at 0 precisely
plot(density_H_neg, main = paste(gene_neg), 
     xlab = "", ylab = "", col = "blue",
     xlim = xlim_neg, ylim = ylim_neg, lwd = 4, cex.axis = 2,
     xaxs = "i", yaxs = "i")

# Add red dashed line and points
abline(v = expression_D_neg, col = "red", lwd = 4, lty = 2)
points(expression_D_neg, 0, col = "red", pch = 19)

```
The blue line indicates the reference gene expression distribution with the red line
indicating the gene expression of the query. Since we only had a very small 
reference for demonstration purposes, the results are not very robust, but can 
show some trends. Plotting example driver genes can help us to better understand 
the underlying gene expression changes. Next, we will repeat the analysis after 
adjusting some CellSpectra parameters.


## Impact of gene filtering
Sparsity is one of the major issues in single-cell analysis. While analysis pipelines
already filter genes of low expression, some genes might be low expressed in 
a specific cell type or in a specific set of samples. Therefore, we can directly
remove genes of low expression in the query or reference gene expression matrix 
before running CellSpectra. The need for filtering genes depends on the cell type, 
cell numbers, the gene set database (number of genes per gene set) or tissu of
investigation. For example, we applied gene filtering to the KEGG analysis in our
manuscript to improve the robustness of our results. However, especially in rare 
cell types zero inflation is very common after pseudobulking, and filtering too 
stringent means that analysis cannot be performed for most pathways. We always 
recommend to apply gene filtering when focusing on driver genes. 

```{r Run Spectra with gene filtering, echo=TRUE, message=FALSE, warning=FALSE}
#Step 2
run_spectra(output_folder_base = output_folder_base, cell_types = c("PT"), CHISQ.MAX = 4, expression_threshold = 0, gene_number_threshold = 10, QC_report = TRUE)

#load results
pval_df = read.csv("output/Padj/Padj_PT.csv", row.names = 1)
pval_df = t(pval_df)

#load results
r2_df = read.csv("output/R2/R2_PT.csv", row.names = 1)
r2_df = t(r2_df)

qc_report_query = read.csv("output/PT/subfolder_sample_32-2/pathway_QC.csv", row.names = 1)
```
Here we see that not filtering genes of zero expression increased the 
number of significant pathways in our query. Gene filtering is applied based on 
the normalized query gene expression matrix. If the average expression of a gene
across the reference samples is below the threshold, genes are also removed
from the analysis.

In summary: Users should think about the degree of sparsity and frequency of low 
expressed genes after pseudobulking in their data, since  a high number of genes 
of zero / low expression, can inflate the statistics. If gene filtering is not applied 
but the data has a high count of zeros / very few cells, it might be advisable to 
check robustness of significant pathways for the filtering of low expressed genes. 
Does my identified pathway of interest stay significant after repeating 
the analysis with a more stringent filtering applied? Can I extract meaningful 
driver genes with significant gene expression changes? Does my QC report indicate
a good estimation of V1, with a small SD? If the answer to all these questions is
no, a critical view on the result might be appropriate. 



## Impact of CHISQ.MAX
We already took a look at our driver genes underlying the "dyscoordination" p value.
One very easy and intuitive interpretation of driver genes in this context is the 
question, if my gene in the query showed a much higher or much lower expression 
level compared to what was expected based on the reference gene expression pattern 
estimation (if our query could be perfectly predicted by the linear model). For 
the chi-square statistic we are applying, we are squaring the "normalized residuals"
that can be extracted with the get_driver_genes() function. However, we don't want
individual genes to drive our p value. Therefore, we define the max. value that 
a gene can have after its "normalized residual" is squared (CHISQ.MAX). 
This parameter does not have an impact on the R2!

Lets check the pvalue after gene filtering for our cAMP Signaling Pathway again: 
padj = 1.365786e-05

Top pos. driver genes and normalized residuals: 
CREB3L2
4.3688108147

CALM3
3.8256698920

RAC1
3.7808482458

A CHISQ.MAX = 4 (CellSpectra's default) corresponds to a normalized residual of 
2 or -2. (2^2 = 4). In other words, any driver gene with a value greater than 
2 is automatically set to 2. This would be the case for all of our top 3 pos 
driver genes. These caveats ensure that the entire pathway is dyscoordinated, 
and not just very few individual genes. 

We will now check what impact CHISQ.MAX = 4 will have on our pval:

```{r Change CHISQ.MAX, echo=TRUE, message=FALSE, warning=FALSE}
#Step 2
run_spectra(output_folder_base = output_folder_base, cell_types = c("PT"), CHISQ.MAX = 3, expression_threshold = 0, gene_number_threshold = 10, QC_report = TRUE)

#load results
pval_df = read.csv("output/Padj/Padj_PT.csv", row.names = 1)
pval_df = t(pval_df)

#load results
r2_df = read.csv("output/R2/R2_PT.csv", row.names = 1)
r2_df = t(r2_df)

qc_report_query = read.csv("output/PT/subfolder_sample_32-2/pathway_QC.csv", row.names = 1)
```
Our new results: 
cAMP.signaling.pathway padj = 4.761164e-03

The lower we set CHISQ.MAX, the more stringent we expect the entire pathway to 
be dyscoordinated. How much individual genes are allowed to drive the measured
dyscoordination depends on the investigators research questions, the gene set 
database, the cell type and sparsity, etc. We recommend to start with the default
setting and to tune the parameters if needed (see below).


## Summary and General Recommendations:
We recommend to start with out default settings: 
CHISQ.MAX = 4, expression_threshold = 0 

Some key questions we should ask after browsing our first results: 

1.) Make sure that the results align with what you would expect from your biological 
system. For instance: Are you able to distinguish healthy and disease based on the
dyscoordination quantity? Are expected pathways represented in the results? 

2.) Check the QC report of you query: How well did the reference set of samples 
estimate V1? Are my reference samples too heterogeneous and noisy (trend towards
low average R2 and high SD, low V1VE)? Is my reference too homogeneous (ratio of
biosamples / pseudoreplicates too small? does my reference really represent the
full spectrum of "normal" or is my reference sample size too small..)? 

3.) Do I have enough statistical power to extract meaningful pathways from my biological 
system or do I see less significant features than what was expected? Is my gene set 
database used in the unsupervised analysis too large (correction for multiple testing)?
Are too many pathways removed from the analysis due to the gene filtering?

4.) Driver genes: Once my significant pathways were identified, can I describe the 
changes of the underlying driver genes and can I confirm these changes in 
other samples of the same condition?


## Pro and Cons of tuning CellSpectra parameters:

I Gene filtering
Pro: 
+ Increases robustness of results
+ Highly recommended for the identification of driver genes

Con:
- Some pathways might be analyzed in one query but not the other 
  (total pathway numbers cannot be compared)
- For sparse cell types many pathways of potential interest might be removed
  from the analysis
- Trends towards less significant features 


II CHISQ.MAX
Pro ( = Decrease):
+ Increases robustness of results 
+ Individual genes cannot drive the dyscoordination
 
Con: 
- Less power for detecting changes
- Some changes might be driven by relatively few genes, 
  but can still be biologically relevant 


III Pseudoreplicates
Pro:
+ Was designed for comparing biopsies of small cell numbers to references with 
  larger cell numbers (considering more cells for the estimation of V1)
+ Adds the factor of sampling bias to the reference
+ Increases the power of the reference in cases of relatively few refsamples 
  but relatively large cell numbers
  
Con: 
- Can introduce false homogeneity within the reference 
  (inflates the number of significant results)
- Especially in cases of relatively small reference samples compared to the query
  or a very small number of reference samples


